---
title: "Final Project EDA"
author: "Akwellé (Q) Quaye"
date: '2022-07-11'
output: html_document
---

#Load county rankings data

```{r}
library(tidyverse)
rankings_data <- read_csv("//Users/Akwelle/Desktop/SURE 22/surefinalproject/RankedMeasureData.csv", col_names = TRUE)
```


#Cleaning county rankings data

```{r}
#removing confidence interval information
rankings_data <- rankings_data %>%
  select(-contains("CI"))

#removing quartile information
rankings_data <- rankings_data %>%
  select(-contains("Quartile"))

#removing counties with unreliable data
rankings_data <- rankings_data %>%
  filter(is.na(Unreliable))

#removing "unreliable" column
rankings_data <- rankings_data %>%
  select(-contains("Unreliable"))

#changing FIPS to a categorical variable
rankings_data$FIPS <- as.factor(rankings_data$FIPS)

#Editing column names for "Population" to match accordingly
colnames(rankings_data)[57] <- "Adults 25 and Over"
colnames(rankings_data)[60] <- "Adults 25-44"

```

# PURPOSE: EDA for county rankings data--------

#scatterplot of % completed high school v. % excessive drinking

```{r}
head(rankings_data)

rankings_data %>%
  ggplot(aes(x = `% Completed High School`, y=`% Excessive Drinking`)) +
  geom_point(alpha = 0.3) +
  theme_minimal()
  
```

#scatterplot of mental health provider rate and % of excessive drinking

```{r}
head(rankings_data)

rankings_data %>%
  ggplot(aes(x = `Mental Health Provider Rate`, y=log(`% Excessive Drinking`))) +
  geom_point(alpha = 0.3) +
  theme_minimal()
```
# scatterplot of % severe housing problems v. % excessive drinking

```{r}
head(rankings_data)

rankings_data %>%
  ggplot(aes(x = `% Severe Housing Problems`, y=`% Excessive Drinking`)) +
  geom_point(alpha = 0.3) +
  theme_minimal()
```


# PURPOSE: PCA to determine which variables are significant to determining % of alcohol-involved driving deaths ============

# Step 1: Removing categorical and NA data

```{r}
#removing categorical data
rankings_numeric <- rankings_data %>%
  select(-c(FIPS, State, County))

# removing ratios data
rankings_numeric <- rankings_numeric %>%
  select(-`Mental Health Provider Ratio` & -`Dentist Ratio`)

#transforming yes/no into 1 and 0
rankings_numeric$`Presence of Water Violation` <- ifelse(rankings_numeric$`Presence of Water Violation` == "Yes", 1, 0)

#removing race information (not relevant to PCA)
rankings_numeric_clean <- rankings_numeric %>%
  select(-contains("Asian") & -contains("AIAN") & -contains("Hispanic") & -contains("Black") & -contains("white"))


```

# Step 2: Run PCA

```{r}

rankings_numeric_clean <- rankings_numeric_clean %>% 
  mutate_all(~ifelse(is.na(.x), mean(.x, na.rm = TRUE), .x))  


### standard regression ###
simple = lm(`# Alcohol-Impaired Driving Deaths` ~ ., data = rankings_numeric_clean)
summary(simple)

plot(x = rankings_numeric_clean$`% Excessive Drinking`,
     y = rankings_numeric_clean$`# Alcohol-Impaired Driving Deaths`)

plot(x = rankings_numeric_clean$`% Driving Deaths with Alcohol Involvement`,
     y = rankings_numeric_clean$`# Alcohol-Impaired Driving Deaths`)

plot(x = rankings_numeric_clean$`# Driving Deaths`,
     y = rankings_numeric_clean$`# Alcohol-Impaired Driving Deaths`)

ix = sort(rankings_numeric_clean$`# Driving Deaths`, decreasing=TRUE,
          index.return = TRUE)$ix[-c(1:3)]
plot(x = rankings_numeric_clean$`# Driving Deaths`[ix],
     y = rankings_numeric_clean$`# Alcohol-Impaired Driving Deaths`[ix])

single = lm(`# Alcohol-Impaired Driving Deaths` ~ `# Driving Deaths`, data = rankings_numeric_clean)
summary(single)

hist(rankings_numeric_clean$`# Alcohol-Impaired Driving Deaths`/
     rankings_numeric_clean$`# Driving Deaths`)

```

```{r}

### pca ###
rankings_pca <- prcomp(rankings_numeric_clean, center = TRUE, scale=TRUE)
summary(rankings_pca)



#####
rankings_numeric_clean <- rankings_numeric_clean %>% 
  mutate_all(~ifelse(is.na(.x), mean(.x, na.rm = TRUE), .x))  


ix = which(colnames(rankings_numeric_clean) == "# Alcohol-Impaired Driving Deaths")
x = rankings_numeric_clean[,-ix]
rankings_pca <- prcomp(x, center = TRUE, scale=TRUE)
summary(rankings_pca)

pca_x = data.frame(rankings_pca$x)
pca_data = pca_x
pca_data$`Alcohol_Impaired_Driving_Deaths` = rankings_numeric_clean$`# Alcohol-Impaired Driving Deaths`
full = lm(Alcohol_Impaired_Driving_Deaths ~ ., data = pca_data)
summary(full)
bic_result <- step(full, trace  =FALSE,
     direction = "backward", criterion = "BIC", k = log(dim(pca_data)[1]))



```

#Step 3: Analyze PCA

```{r}
library(factoextra)


#scree plot 
fviz_eig(rankings_pca) #based off of this, I've determined we only need the first three principal components

#variables
fviz_pca_var(rankings_pca) #note that two PCAs are responsible for 57.2% of the variance

#eigenvalues
eig_rankings <- get_eigenvalue(rankings_pca)
eig_rankings #based on this, the first 9 PCAs should be used
```
# Analysis of PC 1

```{r}
pc1_contribution <- fviz_contrib(rankings_pca, "var", axes=1, xtickslab.rt=45)

plot(pc1_contribution)
```

# Analysis of PC 2
```{r}
pc2_contribution <- fviz_contrib(rankings_pca, "var", axes=2, xtickslab.rt=45)

plot(pc2_contribution)
```

# Analysis of PC 3
```{r}
pc3_contribution <- fviz_contrib(rankings_pca, "var", axes=3, xtickslab.rt=45)

plot(pc3_contribution)
```



# Step 4: Clustering

```{r}
#silhouette analysis to determine the # of clusters

fviz_nbclust(rankings_pca$x, FUNcluster = kmeans, k.max=10)

#conducting clustering

eclust(rankings_pca$x, "kmeans", hc_metric="euclidean", k=3)
```

# Step 5: PCR

```{r}
library(pls)
set.seed(2000)

rankings_pcr = pcr(`# Alcohol-Impaired Driving Deaths`~., data = rankings_numeric_clean, scale=TRUE, validation = "CV")



#validation plot
validationplot(rankings_pcr, val.type = "MSEP")

#split data into train and test sets of data 

rankings_train <- rankings_numeric_clean()
```



# Step 6: Lasso Regression
```{r}
set.seed(2000)
library(glmnet)

model_x <- rankings_numeric_clean %>%
  select(-``)
  

fit_lasso_cv <- cv.glmnet(model_x, model_y, 
                          alpha = 1)
tidy_lasso_coef <- tidy(fit_lasso_cv$glmnet.fit)
tidy_lasso_coef %>%
  ggplot(aes(x = lambda, y = estimate, 
             group = term)) +
  scale_x_log10() +
  geom_line(alpha = 0.75) +
  geom_vline(xintercept = 
               fit_lasso_cv$lambda.min) +
  geom_vline(xintercept = 
               fit_lasso_cv$lambda.1se, 
             linetype = "dashed", color = "red") +
  theme_bw()

```